---
pagetitle: "Chapter 5: Confidence Intervals - Part II"
engine: knitr
format:
  live-revealjs:
    theme: simple
    transition: slide
    slide-number: true
    toc: false
    chalkboard: true
    footer: "STA258 | Chapter 5"
    pdf-separate-fragments: false
    html-math-method: mathjax
    code-copy: true
    code-fold: false
    incremental: false
    width: 1280
    height: 720
    margin: 0.055
    center: false
revealjs-plugins:
  - quiz
webr:
  packages:
    - ggplot2
    - dplyr
live:
  show-solutions: true
editor:
  markdown:
    wrap: none
---

```{r setup}
#| include: false

library(ggplot2)
library(dplyr)
library(plotly)

theme_set(theme_minimal(base_size = 13))
set.seed(258)

plot_cols <- list(
  main          = "#1f4e79",
  main_dark     = "#16324f",
  secondary     = "#2563eb",
  accent        = "#d97706",
  red           = "#b91c1c",
  green         = "#047857",
  purple        = "#6d28d9",
  slate         = "#475569",
  text          = "#111827",
  muted         = "#4b5563",
  grid          = "#e5e7eb",
  border        = "#cbd5e1",
  shade_blue    = "rgba(37, 99, 235, 0.16)",
  shade_orange  = "rgba(217, 119, 6, 0.22)",
  shade_grey    = "rgba(100, 116, 139, 0.14)",
  shade_green   = "rgba(4, 120, 87, 0.14)"
)

plotly_slide_layout <- function(
    p,
    title,
    x_title,
    y_title,
    height = 380,
    x_range = NULL,
    y_range = NULL,
    shapes = list(),
    annotations = list(),
    showlegend = FALSE,
    legend_y = -0.18,
    margin = list(l = 76, r = 28, t = 68, b = 70)
) {
  p |>
    layout(
      title = list(
        text = paste0("<b>", title, "</b>"),
        x = 0.015,
        xanchor = "left",
        font = list(
          size = 20,
          family = "Arial, sans-serif",
          color = plot_cols$text
        )
      ),
      font = list(
        family = "Arial, sans-serif",
        color = plot_cols$text
      ),
      xaxis = list(
        title = list(
          text = paste0("<b>", x_title, "</b>"),
          standoff = 16,
          font = list(size = 15, color = plot_cols$text)
        ),
        range = x_range,
        tickfont = list(size = 13, color = plot_cols$muted),
        showline = TRUE,
        linewidth = 1.2,
        linecolor = plot_cols$border,
        zeroline = FALSE,
        showgrid = FALSE,
        ticks = "outside",
        ticklen = 5,
        tickcolor = plot_cols$border,
        automargin = TRUE
      ),
      yaxis = list(
        title = list(
          text = paste0("<b>", y_title, "</b>"),
          standoff = 18,
          font = list(size = 15, color = plot_cols$text)
        ),
        range = y_range,
        tickfont = list(size = 13, color = plot_cols$muted),
        showline = TRUE,
        linewidth = 1.2,
        linecolor = plot_cols$border,
        zeroline = FALSE,
        gridcolor = plot_cols$grid,
        gridwidth = 1,
        ticks = "outside",
        ticklen = 5,
        tickcolor = plot_cols$border,
        automargin = TRUE
      ),
      showlegend = showlegend,
      legend = list(
        orientation = "h",
        x = 0,
        y = legend_y,
        bgcolor = "rgba(255,255,255,0)",
        font = list(size = 13, color = plot_cols$text)
      ),
      shapes = shapes,
      annotations = annotations,
      margin = margin,
      paper_bgcolor = "rgba(0,0,0,0)",
      plot_bgcolor = "white",
      hovermode = "closest",
      hoverlabel = list(
        bgcolor = "white",
        bordercolor = plot_cols$main,
        font = list(
          size = 13,
          color = plot_cols$text,
          family = "Arial, sans-serif"
        )
      ),
      height = height
    ) |>
    config(
      displayModeBar = TRUE,
      displaylogo = FALSE,
      responsive = TRUE,
      scrollZoom = FALSE,
      doubleClick = "reset",
      modeBarButtonsToRemove = c(
        "lasso2d",
        "select2d"
      )
    )
}

welch_df <- function(s1, n1, s2, n2) {
  numerator <- (s1^2 / n1 + s2^2 / n2)^2
  denominator <- ((s1^2 / n1)^2 / (n1 - 1)) +
    ((s2^2 / n2)^2 / (n2 - 1))
  numerator / denominator
}

sample_size_t_iter <- function(s, E, conf = 0.95, max_iter = 20) {
  alpha <- 1 - conf
  n_current <- ceiling((qnorm(1 - alpha / 2) * s / E)^2)

  for (i in seq_len(max_iter)) {
    t_star <- qt(1 - alpha / 2, df = n_current - 1)
    n_next <- ceiling((t_star * s / E)^2)

    if (n_next == n_current) {
      break
    }

    n_current <- n_next
  }

  n_current
}
```

## 5.4 Two-Sample Confidence Intervals

```{=html}
<style>
:root {
  --sta-teal:        #00695c;
  --sta-teal-dark:   #004d40;
  --sta-teal-light:  #e0f2f1;
  --sta-blue:        #c8e6f3;
  --sta-blue-soft:   #eef8fb;
  --sta-grey:        #f5f6f7;
  --sta-yellow:      #fff8e1;
  --sta-orange:      #f9a825;
  --sta-text:        #2f3437;
  --sta-muted:       #667174;
  --sta-border:      #d8dee2;
  --sta-code:        #f3f4f5;
}

/* ── Base ─────────────────────────────────────────────── */
.reveal {
  font-family: Arial, sans-serif;
  color: var(--sta-text);
}
.reveal .slides section {
  text-align: left;
  padding: 18px 30px;
  box-sizing: border-box;
}
.reveal h1, .reveal h2, .reveal h3 {
  color: var(--sta-text);
  line-height: 1.1;
}
.reveal h1  { font-size: 1.72em !important; }
.reveal h2  {
  font-size: 1.6em !important;
  border-bottom: 3px solid var(--sta-teal);
  padding-bottom: 4px;
  margin-bottom: 12px;
}
.reveal h3  { font-size: 1.1em !important; margin-bottom: 0.2em; }
.reveal p, .reveal li { font-size: 0.78em; line-height: 1.35; }

/* ── Grid layouts ─────────────────────────────────────── */
.two-col, .two-col-lean, .two-col-wide,
.three-col, .four-grid, .web-grid, .exercise-grid, .one-col {
  display: grid;
  gap: 28px;
  align-items: start;
}
.two-col       { grid-template-columns: 1fr 1fr; }
.two-col-lean  { grid-template-columns: 0.95fr 1.05fr; }
.two-col-wide  { grid-template-columns: 1.08fr 0.92fr; }
.three-col     { grid-template-columns: repeat(3, 1fr); }
.one-col       { grid-template-columns: 1fr; }
.four-grid     { grid-template-columns: repeat(2, 1fr); }
.exercise-grid { grid-template-columns: 340px 1fr; gap: 28px; }
.col           { min-width: 0; }

/* ── Boxes ────────────────────────────────────────────── */
.defbox, .examplebox, .thinkbox, .card,
.warnbox, .taskbox, .notebox, .mini-card, .stat-card {
  padding: 16px 18px;
  border-radius: 10px;
  font-size: 0.74em;
  line-height: 1.28;
  box-sizing: border-box;
  margin-bottom: 8px;
}
.defbox, .card {
  background: linear-gradient(160deg, var(--sta-blue), var(--sta-blue-soft));
  border-left: 7px solid var(--sta-teal);
}
.examplebox, .notebox, .mini-card {
  background: var(--sta-grey);
  border: 1px solid var(--sta-border);
}
.thinkbox, .taskbox {
  background: var(--sta-yellow);
  border-left: 7px solid var(--sta-orange);
}
.warnbox {
  background: #fff3e0;
  border-left: 7px solid #ef6c00;
}
.stat-card {
  background: white;
  border: 1px solid var(--sta-border);
  box-shadow: 0 1px 4px rgba(0,0,0,0.07);
}
.card .label, .taskbox .label, .notebox .label,
.mini-card .label, .stat-card .label {
  display: block;
  color: var(--sta-teal-dark);
  font-weight: 700;
  margin-bottom: 5px;
}

/* ── Utility ──────────────────────────────────────────── */
.big        { font-size: 1.08em; font-weight: bold; color: var(--sta-teal); }
.center     { text-align: center; }
.small-text { font-size: 0.64em; }
.tight li   { margin-bottom: 0.18em; }
.formula {
  display: inline-block;
  background: white;
  border: 1px solid var(--sta-border);
  border-radius: 8px;
  padding: 5px 10px;
  font-family: "SFMono-Regular", Consolas, monospace;
  font-size: 1.06em;
}

/* ── Tables ───────────────────────────────────────────── */
.reveal table { width: 100%; border-collapse: collapse; font-size: 0.66em; }
.reveal th, .reveal td {
  border: 1px solid var(--sta-border);
  padding: 7px 10px;
  vertical-align: top;
}
.reveal th { background: var(--sta-teal-light); font-weight: bold; }

/* ── Code ─────────────────────────────────────────────── */
.reveal pre {
  width: 100%;
  margin-top: 0.3em;
  margin-bottom: 0.3em;
  border-radius: 8px;
}
.reveal pre code { font-size: 0.75em; max-height: 340px; }
.reveal code {
  background: var(--sta-code);
  border-radius: 4px;
  padding: 1px 4px;
}

/* ── Scrollable slides ────────────────────────────────── */
.reveal .slides section.scrollable {
  overflow-y: auto !important;
  overflow-x: hidden !important;
  max-height: 700px;
}

/* ── Compact ──────────────────────────────────────────── */
.reveal section.compact h2  { font-size: 1.1em !important; }
.reveal section.compact p,
.reveal section.compact li  { font-size: 0.74em; }

/* ── Quiz polish ──────────────────────────────────────── */
.reveal section.quiz-question {
  display: flex !important;
  flex-direction: column !important;
  justify-content: center !important;
  align-items: stretch !important;
  height: 100%;
}
.reveal .quiz-question > * { width: 100%; max-width: 1080px; }
.reveal .quiz-question h2 { font-size: 1.1em !important; margin-bottom: 10px; text-align: center; }
.reveal .quiz-question .callout-note { margin: 4px 0 12px 0 !important; text-align: left; }
.reveal .quiz-question .callout-note .callout-body p,
.reveal .quiz-question .callout-note p { font-size: 0.78em !important; line-height: 1.3; margin: 4px 0; }
.reveal .quiz-question .callout-title { font-size: 0.82em !important; padding: 5px 12px !important; }
.reveal .quiz-question ul { margin: 4px 0 !important; padding-left: 0 !important; list-style: none !important; }
.reveal .quiz-question li { margin-bottom: 6px !important; }

.reveal .option-button {
  padding: 9px 16px !important;
  margin-bottom: 7px !important;
  font-size: 0.82em !important;
  line-height: 1.25 !important;
  border-radius: 8px !important;
  text-align: left !important;
}
.reveal .action-buttons {
  margin: 10px 0 0 0 !important;
  font-size: 0.85em !important;
  text-align: right;
}
.reveal .check-button,
.reveal .reset-button,
.reveal .prev-button,
.reveal .next-button {
  padding: 7px 16px !important;
  font-size: 0.78em !important;
  border-radius: 6px !important;
  margin: 0 4px !important;
}

.reveal .feedback,
.reveal .explanation,
.reveal .quiz-question .explanation { font-size: 0.7em !important; margin-top: 8px; }

.quiz-hint {
  font-size: 0.6em;
  color: var(--sta-muted);
  font-style: italic;
  margin: 0 0 6px 0;
}

/* ── Plotly polish ───────────────────────────────────── */
.reveal .html-widget,
.reveal .plotly,
.reveal .js-plotly-plot {
  margin: 0 auto !important;
  max-width: 100% !important;
}

.reveal .plot-container {
  border-radius: 10px;
}

.reveal section.quiz-question .html-widget {
  max-height: 315px !important;
  margin: 0 auto 4px auto !important;
}

/* ── Solution callouts ────────────────────────────────── */
.callout { font-size: 0.75em; }
</style>
```

```{=html}
<script>
function resizePlotlyOnSlides() {
  if (!window.Plotly) return;

  document.querySelectorAll(".js-plotly-plot").forEach(function(plot) {
    Plotly.Plots.resize(plot);
  });
}

document.addEventListener("DOMContentLoaded", function () {
  setTimeout(resizePlotlyOnSlides, 250);
  setTimeout(resizePlotlyOnSlides, 750);
  setTimeout(resizePlotlyOnSlides, 1500);

  if (window.Reveal) {
    Reveal.on("ready", function () {
      setTimeout(resizePlotlyOnSlides, 300);
    });

    Reveal.on("slidechanged", function () {
      setTimeout(resizePlotlyOnSlides, 300);
    });

    Reveal.on("fragmentshown", function () {
      setTimeout(resizePlotlyOnSlides, 300);
    });

    Reveal.on("fragmenthidden", function () {
      setTimeout(resizePlotlyOnSlides, 300);
    });
  }

  window.addEventListener("resize", resizePlotlyOnSlides);
});
</script>
```

{{< include ../_extensions/r-wasm/live/_knitr.qmd >}}

```{webr}
#| autorun: true
#| echo: false
#| output: false

welch_df <- function(s1, n1, s2, n2) {
  numerator <- (s1^2 / n1 + s2^2 / n2)^2
  denominator <- ((s1^2 / n1)^2 / (n1 - 1)) +
    ((s2^2 / n2)^2 / (n2 - 1))
  numerator / denominator
}

sample_size_t_iter <- function(s, E, conf = 0.95, max_iter = 20) {
  alpha <- 1 - conf
  n_current <- ceiling((qnorm(1 - alpha / 2) * s / E)^2)

  for (i in seq_len(max_iter)) {
    t_star <- qt(1 - alpha / 2, df = n_current - 1)
    n_next <- ceiling((t_star * s / E)^2)

    if (n_next == n_current) {
      break
    }

    n_current <- n_next
  }

  n_current
}
```

::: {.defbox .fragment}
A **two-sample confidence interval** estimates a population quantity that
compares two groups.

The groups must be clearly labelled, because the order matters:

$$
\text{Group 1} - \text{Group 2}
$$

Changing the order changes the sign of the interval.
:::

:::::: three-col
::: {.stat-card .fragment}
[Difference of means]{.label}

$$
\mu_1-\mu_2
$$

Used for comparing two population averages.
:::

::: {.stat-card .fragment}
[Difference of proportions]{.label}

$$
p_1-p_2
$$

Used for comparing two population percentages.
:::

::: {.stat-card .fragment}
[Ratio of variances]{.label}

$$
\frac{\sigma_1^2}{\sigma_2^2}
$$

Used for comparing two population variances.
:::
::::::

------------------------------------------------------------------------

## 5.4.1 Confidence Interval for a Difference of Means

::: {.defbox .fragment}
For two independent populations, we often want to estimate:

$$
\mu_1-\mu_2
$$

A natural statistic is:

$$
\bar{x}_1-\bar{x}_2
$$
:::

:::::: three-col
::: {.stat-card .fragment}
[Case 1]{.label} $\sigma_1$ and $\sigma_2$ known

Use a **$z$ interval**.

Rare in practice.
:::

::: {.stat-card .fragment}
[Case 2]{.label} $\sigma_1$ and $\sigma_2$ unknown, but assumed equal

Use a **pooled two-sample $t$ interval**.
:::

::: {.stat-card .fragment}
[Case 3]{.label} $\sigma_1$ and $\sigma_2$ unknown and not assumed equal

Use **Welch's two-sample $t$ interval**.
:::
::::::

------------------------------------------------------------------------

## Difference of Means: Known $\sigma_1$ and $\sigma_2$

::: defbox
When the population standard deviations are known, a level
$100(1-\alpha)\%$ confidence interval for $\mu_1-\mu_2$ is:

$$
(\bar{x}_1-\bar{x}_2)
\pm
z_{1-\alpha/2}
\sqrt{
\frac{\sigma_1^2}{n_1}
+
\frac{\sigma_2^2}{n_2}
}
$$
:::

::: {.warnbox .fragment}
This case is mathematically useful but rare in real applications, because
population standard deviations are usually unknown.
:::

------------------------------------------------------------------------

## Example 1: Known $\sigma_1$ and $\sigma_2$ {.scrollable}

:::::: two-col-wide
:::: col
::: taskbox
[Example]{.label} Suppose final marks in the same course are compared
between two campuses.

Group 1: UTSG  
$n_1=64$, $\bar{x}_1=76$, known $\sigma_1=10$

Group 2: UTM  
$n_2=81$, $\bar{x}_2=72$, known $\sigma_2=12$

Construct a 95% confidence interval for $\mu_1-\mu_2$.

**(a)** What is the point estimate?

**(b)** What is the standard error?

**(c)** Construct and interpret the interval.
:::
::::

::: col
::: {.examplebox}
**Set-up**

$$
\bar{x}_1-\bar{x}_2 = 76-72 = 4
$$

Because both population standard deviations are known, this is the rare
two-sample **$z$ interval** case.

The interval estimates:

$$
\mu_{\text{UTSG}}-\mu_{\text{UTM}}
$$
:::

::: {.thinkbox}
A positive interval would suggest the UTSG population mean is larger. An
interval containing 0 would mean 0 is a plausible difference.
:::
:::
::::::

------------------------------------------------------------------------

## Known $\sigma_1,\sigma_2$ Example — Check in R

```{webr}
#| autorun: false

xbar1 <- 76
xbar2 <- 72
sigma1 <- 10
sigma2 <- 12
n1 <- 64
n2 <- 81
conf <- 0.95

alpha <- 1 - conf
z_star <- qnorm(1 - alpha / 2)

estimate <- xbar1 - xbar2
se <- sqrt(sigma1^2 / n1 + sigma2^2 / n2)
me <- z_star * se

lower <- estimate - me
upper <- estimate + me

c(
  estimate = estimate,
  standard_error = se,
  margin_of_error = me,
  lower = lower,
  upper = upper
)
```

::: {.small-text}
Using the values above, the 95% confidence interval is approximately
$(0.42,\; 7.58)$.
:::

------------------------------------------------------------------------

## Difference of Means: Equal Unknown Variances

::: defbox
If $\sigma_1$ and $\sigma_2$ are unknown but assumed equal, we pool the
two sample variances.

The pooled variance is:

$$
s_p^2 =
\frac{(n_1-1)s_1^2+(n_2-1)s_2^2}{n_1+n_2-2}
$$

The level $100(1-\alpha)\%$ confidence interval for $\mu_1-\mu_2$ is:

$$
(\bar{x}_1-\bar{x}_2)
\pm
t_{1-\alpha/2,\;n_1+n_2-2}
s_p
\sqrt{\frac{1}{n_1}+\frac{1}{n_2}}
$$
:::

::: {.thinkbox .fragment}
Use this only when the equal-variance assumption is reasonable.
Otherwise, use Welch's interval.
:::

------------------------------------------------------------------------

## Example 1: Pooled Two-Sample $t$ Interval {.scrollable}

:::::: two-col-wide
:::: col
::: taskbox
[Example]{.label} A researcher compares battery life for two brands of
the same device.

Brand A: $n_1=12$, $\bar{x}_1=52.4$, $s_1=6.1$  
Brand B: $n_2=10$, $\bar{x}_2=47.8$, $s_2=5.4$

Assume the two population variances are equal.

Construct a 95% confidence interval for $\mu_1-\mu_2$.

**(a)** Why is this a pooled two-sample interval?

**(b)** Calculate the pooled standard deviation.

**(c)** Construct and interpret the confidence interval.
:::
::::

::: col
::: {.examplebox}
**Parameter**

$$
\mu_1-\mu_2
$$

where group 1 is Brand A and group 2 is Brand B.

**Point estimate**

$$
\bar{x}_1-\bar{x}_2 = 52.4-47.8 = 4.6
$$
:::

::: {.thinkbox}
The interval estimates how much longer Brand A lasts on average compared
with Brand B.
:::
:::
::::::

------------------------------------------------------------------------

## Pooled Example 1 — Check in R

```{webr}
#| autorun: false

n1 <- 12
xbar1 <- 52.4
s1 <- 6.1

n2 <- 10
xbar2 <- 47.8
s2 <- 5.4

conf <- 0.95
alpha <- 1 - conf
df <- n1 + n2 - 2

sp2 <- ((n1 - 1) * s1^2 + (n2 - 1) * s2^2) / df
sp <- sqrt(sp2)

estimate <- xbar1 - xbar2
se <- sp * sqrt(1 / n1 + 1 / n2)
t_star <- qt(1 - alpha / 2, df = df)
me <- t_star * se

lower <- estimate - me
upper <- estimate + me

c(
  pooled_sd = sp,
  df = df,
  estimate = estimate,
  standard_error = se,
  margin_of_error = me,
  lower = lower,
  upper = upper
)
```

::: {.small-text}
The 95% confidence interval is approximately $(-0.58,\; 9.78)$.
Because the interval contains 0, the data do not clearly show a nonzero
mean difference at this confidence level.
:::

------------------------------------------------------------------------

## Example 2: Pooled Two-Sample $t$ Interval {.scrollable}

:::::: two-col-wide
:::: col
::: taskbox
[Example]{.label} Two tutorial methods are compared using final quiz
scores.

Method A: $n_1=15$, $\bar{x}_1=84.2$, $s_1=7.5$  
Method B: $n_2=14$, $\bar{x}_2=78.6$, $s_2=6.8$

Assume the two population variances are equal.

Construct a 90% confidence interval for $\mu_1-\mu_2$.

**(a)** State the degrees of freedom.

**(b)** Compute the pooled standard deviation.

**(c)** Interpret the interval in context.
:::
::::

::: col
::: {.examplebox}
**Parameter**

$$
\mu_{\text{Method A}}-\mu_{\text{Method B}}
$$

**Confidence level**

$$
90\%
$$

This means $\alpha = 0.10$ and we use the 95th percentile of the
$t$ distribution because the interval has two tails.
:::

::: {.thinkbox}
The equal-variance assumption lets us combine information from both
sample standard deviations into one pooled standard deviation.
:::
:::
::::::

------------------------------------------------------------------------

## Pooled Example 2 — Check in R

```{webr}
#| autorun: false

n1 <- 15
xbar1 <- 84.2
s1 <- 7.5

n2 <- 14
xbar2 <- 78.6
s2 <- 6.8

conf <- 0.90
alpha <- 1 - conf
df <- n1 + n2 - 2

sp2 <- ((n1 - 1) * s1^2 + (n2 - 1) * s2^2) / df
sp <- sqrt(sp2)

estimate <- xbar1 - xbar2
se <- sp * sqrt(1 / n1 + 1 / n2)
t_star <- qt(1 - alpha / 2, df = df)
me <- t_star * se

lower <- estimate - me
upper <- estimate + me

c(
  pooled_sd = sp,
  df = df,
  estimate = estimate,
  standard_error = se,
  margin_of_error = me,
  lower = lower,
  upper = upper
)
```

::: {.small-text}
The 90% confidence interval is approximately $(1.06,\; 10.14)$. Since
the entire interval is above 0, Method A appears to have a higher mean
score.
:::

------------------------------------------------------------------------

## Difference of Means: Unequal Unknown Variances

::: {.defbox}
If the population variances are **unknown** and are **not assumed equal**, use **Welch’s two-sample $t$ interval**:

$$
(\bar{x}_1 - \bar{x}_2)
\pm
t_{1-\alpha/2,\,df}
\sqrt{
\frac{s_1^2}{n_1}
+
\frac{s_2^2}{n_2}
}
$$

This estimates:

$$
\mu_1 - \mu_2
$$

when the two samples are **independent** and the equal-variance assumption is not reasonable.
:::

::: {.thinkbox}
**Course note.** Welch’s method is the more flexible default when the two sample variances look meaningfully different.
:::

---

## Welch Degrees of Freedom

::: {.notebox}
<span class="label">Degrees of freedom note</span>

In this course, we are **not testing your ability to be a calculator** in an exam setting.

For hand calculations, you may use the **conservative degrees of freedom**:

$$
df = \min(n_1 - 1,\; n_2 - 1)
$$
:::

::: {.defbox}
A more accurate approximation, commonly used by software, is the **Welch–Satterthwaite degrees of freedom**:

$$
df =
\frac{
\left(\frac{s_1^2}{n_1}+\frac{s_2^2}{n_2}\right)^2
}{
\frac{\left(s_1^2/n_1\right)^2}{n_1-1}
+
\frac{\left(s_2^2/n_2\right)^2}{n_2-1}
}
$$
:::

::: {.thinkbox}
For exams in this course, use the conservative version unless told otherwise.
:::
------------------------------------------------------------------------

## Example 1: Welch Two-Sample $t$ Interval {.scrollable}

:::::: two-col-wide
:::: col
::: taskbox
[Example]{.label} Suppose commute times are compared between two groups
of students.

Group 1: $n_1=25$, $\bar{x}_1=42$, $s_1=14$  
Group 2: $n_2=30$, $\bar{x}_2=35$, $s_2=9$

Do not assume equal population variances.

Construct a 95% confidence interval for $\mu_1-\mu_2$.

**(a)** Why is Welch's interval appropriate?

**(b)** Compute the standard error.

**(c)** Construct and interpret the interval.
:::
::::

::: col
::: {.examplebox}
**Point estimate**

$$
\bar{x}_1-\bar{x}_2 = 42-35 = 7
$$

The first group has a sample mean commute time 7 minutes higher.
:::

::: {.warnbox}
Do not pool the variances here. The sample standard deviations are quite
different: $14$ and $9$.
:::
:::
::::::

------------------------------------------------------------------------

## Welch Example 1 — Check in R

```{webr}
#| autorun: false

n1 <- 25
xbar1 <- 42
s1 <- 14

n2 <- 30
xbar2 <- 35
s2 <- 9

conf <- 0.95
alpha <- 1 - conf

estimate <- xbar1 - xbar2
se <- sqrt(s1^2 / n1 + s2^2 / n2)
df <- welch_df(s1, n1, s2, n2)

t_star <- qt(1 - alpha / 2, df = df)
me <- t_star * se

lower <- estimate - me
upper <- estimate + me

c(
  df = df,
  estimate = estimate,
  standard_error = se,
  margin_of_error = me,
  lower = lower,
  upper = upper
)
```

::: {.small-text}
The 95% confidence interval is approximately $(0.43,\; 13.57)$. Since
the interval is above 0, Group 1 appears to have a higher mean commute
time.
:::

------------------------------------------------------------------------

## Example 2: Welch Two-Sample $t$ Interval {.scrollable}

:::::: two-col-wide
:::: col
::: taskbox
[Example]{.label} Weekly food spending is compared between two groups of
students.

Group 1: $n_1=18$, $\bar{x}_1=96$, $s_1=22$  
Group 2: $n_2=22$, $\bar{x}_2=82$, $s_2=16$

Do not assume equal population variances.

Construct a 90% confidence interval for $\mu_1-\mu_2$.

**(a)** State the point estimate.

**(b)** Use the Welch degrees of freedom.

**(c)** Interpret the interval.
:::
::::

::: col
::: {.examplebox}
**Parameter**

$$
\mu_1-\mu_2
$$

where group 1 is the first student group and group 2 is the second
student group.

**Point estimate**

$$
96-82=14
$$
:::

::: {.thinkbox}
The confidence interval tells us a plausible range for the true
difference in average weekly spending.
:::
:::
::::::

------------------------------------------------------------------------

## Welch Example 2 — Check in R

```{webr}
#| autorun: false

n1 <- 18
xbar1 <- 96
s1 <- 22

n2 <- 22
xbar2 <- 82
s2 <- 16

conf <- 0.90
alpha <- 1 - conf

estimate <- xbar1 - xbar2
se <- sqrt(s1^2 / n1 + s2^2 / n2)
df <- welch_df(s1, n1, s2, n2)

t_star <- qt(1 - alpha / 2, df = df)
me <- t_star * se

lower <- estimate - me
upper <- estimate + me

c(
  df = df,
  estimate = estimate,
  standard_error = se,
  margin_of_error = me,
  lower = lower,
  upper = upper
)
```

::: {.small-text}
The 90% confidence interval is approximately $(3.47,\; 24.53)$.
Since the entire interval is above 0, the data suggest that Group 1 has
a higher mean weekly food spending than Group 2.
:::

------------------------------------------------------------------------

## Assumptions for Two-Sample Mean Intervals

:::::: two-col
::: col
::: defbox
**Independent samples**

The two samples must be independent. One person's measurement in group 1
should not determine a person's measurement in group 2.
:::

::: defbox
**Randomization**

Each group should come from a random sample or a randomized experiment.
:::
:::

::: col
::: defbox
**Normality or large samples**

If sample sizes are small, each population should be approximately
normal. For larger samples, the CLT makes the interval more reliable.
:::

::: defbox
**Equal variances only for pooled intervals**

The pooled method requires the extra assumption that
$\sigma_1^2=\sigma_2^2$.
:::
:::
::::::

------------------------------------------------------------------------

## Assumptions for Two-Sample Mean Intervals

:::::: two-col
::: col
::: defbox
**Independent samples**

The two samples must be independent. One person's measurement in group 1
should not determine a person's measurement in group 2.
:::

::: defbox
**Randomization**

Each group should come from a random sample or a randomized experiment.
:::
:::

::: col
::: defbox
**Normality or large samples**

If sample sizes are small, each population should be approximately
normal. For larger samples, the CLT makes the interval more reliable.
:::

::: defbox
**Equal variances only for pooled intervals**

The pooled method requires the extra assumption that
$\sigma_1^2=\sigma_2^2$.
:::
:::
::::::

------------------------------------------------------------------------

##  {.quiz-question}

::: {.callout-note icon="false"}
## Question

Two independent samples are used to compare average commute times. The
population standard deviations are unknown, and the sample standard
deviations are noticeably different. Which interval is the most
appropriate?
:::

-   Known-$\sigma$ two-sample $z$ interval
-   Pooled two-sample $t$ interval
-   [Welch two-sample $t$ interval]{.correct
    data-explanation="Welch's two-sample t interval is appropriate when the two samples are independent, the population standard deviations are unknown, and equal variances are not assumed."}
-   Paired $t$ interval

------------------------------------------------------------------------

## 5.4.2 Confidence Interval for a Difference of Proportions

::: defbox
For two independent samples, let:

$$
\hat{p}_1=\frac{x_1}{n_1},
\qquad
\hat{p}_2=\frac{x_2}{n_2}
$$

A level $100(1-\alpha)\%$ confidence interval for $p_1-p_2$ is:

$$
(\hat{p}_1-\hat{p}_2)
\pm
z_{1-\alpha/2}
\sqrt{
\frac{\hat{p}_1(1-\hat{p}_1)}{n_1}
+
\frac{\hat{p}_2(1-\hat{p}_2)}{n_2}
}
$$
:::

::: {.thinkbox .fragment}
The interval estimates the difference between two population
proportions, not the difference between two sample counts.
:::

------------------------------------------------------------------------

## Example 1: Difference of Proportions {.scrollable}

:::::: two-col-wide
:::: col
::: taskbox
[Example]{.label} Suppose we compare the proportion of left-handed
students at two campuses.

Campus 1: $18$ left-handed students out of $120$  
Campus 2: $10$ left-handed students out of $100$

Construct a 95% confidence interval for $p_1-p_2$.

**(a)** Calculate $\hat{p}_1$ and $\hat{p}_2$.

**(b)** Calculate the standard error.

**(c)** Construct and interpret the interval.
:::
::::

::: col
::: {.examplebox}
**Sample proportions**

$$
\hat{p}_1=\frac{18}{120}=0.15
$$

$$
\hat{p}_2=\frac{10}{100}=0.10
$$

**Point estimate**

$$
\hat{p}_1-\hat{p}_2=0.05
$$
:::

::: {.thinkbox}
The interval estimates how much larger the population proportion is at
Campus 1 compared with Campus 2.
:::
:::
::::::

------------------------------------------------------------------------

## Difference of Proportions Example 1 — Check in R

```{webr}
#| autorun: false

x1 <- 18
n1 <- 120
x2 <- 10
n2 <- 100
conf <- 0.95

p1_hat <- x1 / n1
p2_hat <- x2 / n2

estimate <- p1_hat - p2_hat
se <- sqrt(
  p1_hat * (1 - p1_hat) / n1 +
    p2_hat * (1 - p2_hat) / n2
)

alpha <- 1 - conf
z_star <- qnorm(1 - alpha / 2)
me <- z_star * se

lower <- estimate - me
upper <- estimate + me

c(
  p1_hat = p1_hat,
  p2_hat = p2_hat,
  estimate = estimate,
  standard_error = se,
  margin_of_error = me,
  lower = lower,
  upper = upper
)
```

::: {.small-text}
The 95% confidence interval is approximately $(-0.037,\; 0.137)$.
Since it contains 0, a zero difference is plausible.
:::

------------------------------------------------------------------------

## Example 2: Difference of Proportions {.scrollable}

:::::: two-col-wide
:::: col
::: taskbox
[Example]{.label} A survey compares support for a policy in two cities.

City 1: $210$ support out of $500$  
City 2: $170$ support out of $450$

Construct a 90% confidence interval for $p_1-p_2$.

**(a)** Calculate the two sample proportions.

**(b)** Construct the confidence interval.

**(c)** Does the interval suggest City 1 has a clearly higher support
rate?
:::
::::

::: col
::: {.examplebox}
**Parameter**

$$
p_1-p_2
$$

where $p_1$ is the population support proportion in City 1 and $p_2$ is
the population support proportion in City 2.
:::

::: {.thinkbox}
For a difference of proportions, the standard error uses both sample
proportions.
:::
:::
::::::

------------------------------------------------------------------------

## Difference of Proportions Example 2 — Check in R

```{webr}
#| autorun: false

x1 <- 210
n1 <- 500
x2 <- 170
n2 <- 450
conf <- 0.90

p1_hat <- x1 / n1
p2_hat <- x2 / n2

estimate <- p1_hat - p2_hat
se <- sqrt(
  p1_hat * (1 - p1_hat) / n1 +
    p2_hat * (1 - p2_hat) / n2
)

alpha <- 1 - conf
z_star <- qnorm(1 - alpha / 2)
me <- z_star * se

lower <- estimate - me
upper <- estimate + me

c(
  p1_hat = p1_hat,
  p2_hat = p2_hat,
  estimate = estimate,
  standard_error = se,
  margin_of_error = me,
  lower = lower,
  upper = upper
)
```

::: {.small-text}
The 90% confidence interval is approximately $(-0.010,\; 0.095)$.
Because 0 is inside the interval, the data do not clearly establish a
higher population support rate for City 1.
:::

------------------------------------------------------------------------

## Assumptions for Difference of Proportions

:::::: two-col
::: col
::: defbox
**Randomization**

Each sample should be random, or the data should come from a randomized
experiment.
:::

::: defbox
**Independent groups**

The two groups being compared must be independent from each other.
:::
:::

::: col
::: defbox
**10% condition**

If sampling without replacement, each sample should be less than 10% of
its population.
:::

::: defbox
**Success-failure condition**

Each group should have enough successes and failures:

$$
x_i \ge 10,
\qquad
n_i-x_i \ge 10
$$
:::
:::
::::::

------------------------------------------------------------------------

##  {.quiz-question}

::: {.callout-note icon="false"}
## Question

A two-sample confidence interval for $p_1-p_2$ is constructed. Which
condition checks whether the normal approximation is reasonable?
:::

-   The two sample means must be equal
-   The two population variances must be equal
-   [Each group should have enough successes and failures]{.correct
    data-explanation="For a confidence interval on a difference of proportions, each group should have enough successes and failures for the normal approximation to be reasonable."}
-   The sample sizes must be exactly the same

------------------------------------------------------------------------

## 5.4.3 Confidence Interval for a Ratio of Variances

::: defbox
To compare two population variances, the parameter is:

$$
\frac{\sigma_1^2}{\sigma_2^2}
$$

If both populations are normal and the samples are independent, then:

$$
\frac{s_1^2/\sigma_1^2}{s_2^2/\sigma_2^2}
\sim
F_{n_1-1,\;n_2-1}
$$
:::

::: {.thinkbox .fragment}
A ratio near 1 suggests similar variances. A ratio far above or below 1
suggests one population is more variable.
:::

------------------------------------------------------------------------

## Ratio of Variances: Formula

::: defbox
A level $100(1-\alpha)\%$ confidence interval for
$\sigma_1^2/\sigma_2^2$ is:

$$
\left(
\frac{s_1^2/s_2^2}{F_{1-\alpha/2,\;n_1-1,\;n_2-1}},
\;
\frac{s_1^2/s_2^2}{F_{\alpha/2,\;n_1-1,\;n_2-1}}
\right)
$$
:::

::: {.warnbox .fragment}
This interval is sensitive to the normality assumption. For variance
ratios, normality matters more than it does for large-sample mean
intervals.
:::

------------------------------------------------------------------------

## Example 1: Ratio of Variances {.scrollable}

:::::: two-col-wide
:::: col
::: taskbox
[Example]{.label} Two machines fill bottles. We want to compare the
population variances in fill amount.

Machine 1: $n_1=16$, $s_1=12$  
Machine 2: $n_2=14$, $s_2=8$

Construct a 95% confidence interval for
$\sigma_1^2/\sigma_2^2$.

**(a)** Calculate the sample variance ratio.

**(b)** State the two degrees of freedom.

**(c)** Construct and interpret the interval.
:::
::::

::: col
::: {.examplebox}
**Sample variance ratio**

$$
\frac{s_1^2}{s_2^2}
=
\frac{12^2}{8^2}
=
2.25
$$

The sample variance for Machine 1 is 2.25 times the sample variance for
Machine 2.
:::

::: {.thinkbox}
The confidence interval estimates the ratio of population variances, not
the ratio of sample standard deviations.
:::
:::
::::::

------------------------------------------------------------------------

## Ratio of Variances Example 1 — Check in R

```{webr}
#| autorun: false

n1 <- 16
s1 <- 12

n2 <- 14
s2 <- 8

conf <- 0.95
alpha <- 1 - conf

df1 <- n1 - 1
df2 <- n2 - 1

ratio <- s1^2 / s2^2

lower <- ratio / qf(1 - alpha / 2, df1 = df1, df2 = df2)
upper <- ratio / qf(alpha / 2, df1 = df1, df2 = df2)

c(
  sample_variance_ratio = ratio,
  df1 = df1,
  df2 = df2,
  lower = lower,
  upper = upper
)
```

::: {.small-text}
The 95% confidence interval is approximately $(0.74,\; 6.58)$. Since the
interval contains 1, equal population variances are plausible.
:::

------------------------------------------------------------------------

## Example 2: Ratio of Variances {.scrollable}

:::::: two-col-wide
:::: col
::: taskbox
[Example]{.label} Two labs measure the same chemical concentration using
different procedures.

Lab 1: $n_1=21$, $s_1=5.5$  
Lab 2: $n_2=19$, $s_2=3.8$

Construct a 90% confidence interval for
$\sigma_1^2/\sigma_2^2$.

**(a)** Find the sample variance ratio.

**(b)** Construct the interval using the $F$ distribution.

**(c)** Does the interval suggest clearly different population
variances?
:::
::::

::: col
::: {.examplebox}
**Parameter**

$$
\frac{\sigma_1^2}{\sigma_2^2}
$$

**Interpretation rule**

- If the interval contains 1, equal variances are plausible.
- If the interval is entirely above 1, group 1 appears more variable.
- If the interval is entirely below 1, group 2 appears more variable.
:::
:::
::::::

------------------------------------------------------------------------

## Ratio of Variances Example 2 — Check in R

```{webr}
#| autorun: false

n1 <- 21
s1 <- 5.5

n2 <- 19
s2 <- 3.8

conf <- 0.90
alpha <- 1 - conf

df1 <- n1 - 1
df2 <- n2 - 1

ratio <- s1^2 / s2^2

lower <- ratio / qf(1 - alpha / 2, df1 = df1, df2 = df2)
upper <- ratio / qf(alpha / 2, df1 = df1, df2 = df2)

c(
  sample_variance_ratio = ratio,
  df1 = df1,
  df2 = df2,
  lower = lower,
  upper = upper
)
```

::: {.small-text}
The 90% confidence interval is approximately $(0.96,\; 4.51)$. Since the
interval barely contains 1, equal variances are still plausible at this
confidence level.
:::

------------------------------------------------------------------------

## Assumptions for Ratio of Variances

:::::: two-col
::: col
::: defbox
**Independent samples**

The two samples must be independent from each other.
:::

::: defbox
**Random sampling**

Each sample should be obtained through a random process.
:::
:::

::: col
::: defbox
**Normal populations**

The two populations should be approximately normal. This is especially
important for inference about variances.
:::

::: defbox
**Same measurement scale**

The two variables should be measured on the same or comparable scale.
:::
:::
::::::

------------------------------------------------------------------------

##  {.quiz-question}

::: {.callout-note icon="false"}
## Question

A confidence interval for $\sigma_1^2/\sigma_2^2$ contains the value 1.
What does that mean?
:::

-   The two sample means must be equal
-   [Equal population variances are plausible]{.correct
    data-explanation="The value 1 represents equal variances because sigma_1 squared divided by sigma_2 squared equals 1 when the two population variances are equal."}
-   The two sample sizes must be equal
-   The first variance must be larger than the second variance

------------------------------------------------------------------------

## 5.5 Confidence Intervals on Paired Data

::: defbox
Paired data occur when observations are naturally matched.

Instead of comparing two independent sample means, we compute differences:

$$
d_i = x_{2i} - x_{1i}
$$

Then we build a one-sample $t$ interval for the population mean
difference $\mu_d$.
:::

::::: two-col
::: {.examplebox .fragment}
**Common paired settings**

- Before and after measurements on the same person
- Matched twins
- Same student under two testing conditions
- Same object measured by two devices
:::

::: {.warnbox .fragment}
Do not treat paired data as two independent samples. The pairing is the
main information.
:::
:::::

------------------------------------------------------------------------

## Paired Data: Formula

::: defbox
Let $d_1,d_2,\ldots,d_n$ be the paired differences.

The level $100(1-\alpha)\%$ confidence interval for $\mu_d$ is:

$$
\bar{d}
\pm
t_{1-\alpha/2,\;n-1}
\frac{s_d}{\sqrt{n}}
$$

where $\bar{d}$ is the sample mean of the differences and $s_d$ is the
sample standard deviation of the differences.
:::

::: {.thinkbox .fragment}
The normality condition applies to the **differences**, not necessarily
to the original two measurements separately.
:::

------------------------------------------------------------------------

## Visual: Paired Data Means Differences

```{r}
#| echo: false
#| warning: false
#| message: false
#| out-width: "100%"

student <- paste("Student", 1:10)
before <- c(68, 72, 75, 70, 66, 80, 74, 69, 77, 71)
after <- c(73, 74, 80, 72, 70, 84, 78, 73, 80, 76)

paired_long <- data.frame(
  student = rep(student, each = 2),
  time = rep(c("Before", "After"), times = 10),
  score = as.vector(rbind(before, after))
)

paired_long$time_num <- ifelse(paired_long$time == "Before", 1, 2)

paired_segments <- data.frame(
  student = student,
  before = before,
  after = after
)

p <- plot_ly() |>
  add_segments(
    data = paired_segments,
    x = 1,
    xend = 2,
    y = ~before,
    yend = ~after,
    line = list(color = "rgba(80,95,105,0.45)", width = 2.2),
    hoverinfo = "skip",
    showlegend = FALSE
  ) |>
  add_markers(
    data = paired_long,
    x = ~time_num,
    y = ~score,
    marker = list(
      color = plot_cols$main,
      size = 8,
      line = list(color = "white", width = 1.2)
    ),
    text = ~paste0(
      "<b>", student, "</b>",
      "<br>", time, " score: ", score
    ),
    hovertemplate = "%{text}<extra></extra>",
    showlegend = FALSE
  )

plotly_slide_layout(
  p,
  title = "Paired observations compare within each student",
  x_title = "",
  y_title = "Score",
  height = 390,
  x_range = c(0.75, 2.25),
  y_range = c(64, 86),
  margin = list(l = 70, r = 25, t = 70, b = 60)
) |>
  layout(
    xaxis = list(
      tickvals = c(1, 2),
      ticktext = c("Before", "After"),
      showgrid = FALSE,
      showline = TRUE,
      linecolor = plot_cols$border,
      zeroline = FALSE,
      tickfont = list(size = 13)
    ),
    yaxis = list(
      showgrid = TRUE,
      gridcolor = "rgba(0,0,0,0.08)",
      showline = TRUE,
      linecolor = plot_cols$border,
      zeroline = FALSE,
      tickfont = list(size = 12)
    )
  )
```

::: {.thinkbox}
Each line represents one student. The paired analysis uses the within
student change, not two unrelated groups.
:::

------------------------------------------------------------------------

## Example 1: Paired Confidence Interval {.scrollable}

:::::: two-col-wide
:::: col
::: taskbox
[Example]{.label} Ten students take a quiz before and after a tutoring
session.

The differences are defined as:

$$
d_i = \text{after}_i-\text{before}_i
$$

The observed differences are:

$$
5,\;2,\;5,\;2,\;4,\;4,\;4,\;4,\;3,\;5
$$

Construct a 95% confidence interval for the mean improvement $\mu_d$.

**(a)** Why is this paired data?

**(b)** Find $\bar{d}$ and $s_d$.

**(c)** Construct and interpret the interval.
:::
::::

::: col
::: {.examplebox}
Because the same students are measured before and after tutoring, the
observations are paired.

We analyze the differences:

$$
d_i = \text{after}_i-\text{before}_i
$$

not the two raw score columns as independent samples.
:::
:::
::::::

------------------------------------------------------------------------

## Paired Example 1 — Check in R

```{webr}
#| autorun: false

d <- c(5, 2, 5, 2, 4, 4, 4, 4, 3, 5)

n <- length(d)
dbar <- mean(d)
sd_d <- sd(d)

conf <- 0.95
alpha <- 1 - conf
t_star <- qt(1 - alpha / 2, df = n - 1)

se <- sd_d / sqrt(n)
me <- t_star * se

lower <- dbar - me
upper <- dbar + me

c(
  n = n,
  mean_difference = dbar,
  sd_difference = sd_d,
  standard_error = se,
  margin_of_error = me,
  lower = lower,
  upper = upper
)

t.test(d, conf.level = 0.95)
```

::: {.small-text}
The 95% confidence interval is approximately $(2.99,\; 4.61)$. The
interval is entirely positive, suggesting an average improvement.
:::

------------------------------------------------------------------------

## Example 2: Paired Confidence Interval {.scrollable}

:::::: two-col-wide
:::: col
::: taskbox
[Example]{.label} A medication study measures blood pressure before and
after treatment for 12 patients.

Define the reduction as:

$$
d_i = \text{before}_i-\text{after}_i
$$

The observed reductions are:

$$
6,\;8,\;3,\;10,\;7,\;5,\;9,\;4,\;6,\;8,\;7,\;5
$$

Construct a 90% confidence interval for the mean reduction $\mu_d$.

**(a)** What is the parameter?

**(b)** Why is this not a two-independent-sample problem?

**(c)** Construct and interpret the interval.
:::
::::

::: col
::: {.examplebox}
**Parameter**

$$
\mu_d
$$

where $d_i$ is the blood pressure reduction for patient $i$.

A positive mean difference means the treatment lowers blood pressure on
average.
:::

::: {.thinkbox}
The pairing controls for patient-to-patient differences because each
patient is compared with themselves.
:::
:::
::::::

------------------------------------------------------------------------

## Paired Example 2 — Check in R

```{webr}
#| autorun: false

d <- c(6, 8, 3, 10, 7, 5, 9, 4, 6, 8, 7, 5)

n <- length(d)
dbar <- mean(d)
sd_d <- sd(d)

conf <- 0.90
alpha <- 1 - conf
t_star <- qt(1 - alpha / 2, df = n - 1)

se <- sd_d / sqrt(n)
me <- t_star * se

lower <- dbar - me
upper <- dbar + me

c(
  n = n,
  mean_difference = dbar,
  sd_difference = sd_d,
  standard_error = se,
  margin_of_error = me,
  lower = lower,
  upper = upper
)

t.test(d, conf.level = 0.90)
```

::: {.small-text}
The 90% confidence interval is approximately $(5.43,\; 7.57)$.
:::

------------------------------------------------------------------------

## Assumptions for Paired Data

:::::: two-col
::: col
::: defbox
**Paired observations**

Each pair must come from the same unit or from meaningfully matched
units.
:::

::: defbox
**Independent pairs**

The pairs themselves should be independent of each other.
:::
:::

::: col
::: defbox
**Normal differences**

The distribution of differences should be approximately normal,
especially for small samples.
:::

::: defbox
**Random sampling**

The sample of pairs should come from a random process or a randomized
design.
:::
:::
::::::

------------------------------------------------------------------------

##  {.quiz-question}

::: {.callout-note icon="false"}
## Question

A group of students takes a test before and after a review session. Which
quantity should be used to build the confidence interval?
:::

-   The mean of the before scores only
-   The mean of the after scores only
-   [The mean of the paired differences]{.correct
    data-explanation="Because the same students are measured twice, the data are paired. We compute one difference for each student and build a one-sample t interval for the mean difference."}
-   The ratio of the two sample variances

------------------------------------------------------------------------

## 5.6 Sample Size Selection Using Confidence Intervals

::: defbox
Sample size selection works backwards.

Instead of asking:

> What margin of error do we get from this sample size?

we ask:

> What sample size do we need to achieve this margin of error?
:::

::::: two-col
::: {.examplebox .fragment}
**For a mean**

The required sample size depends on:

- confidence level
- desired margin of error
- population standard deviation, or an estimate of it
:::

::: {.examplebox .fragment}
**For a proportion**

The required sample size depends on:

- confidence level
- desired margin of error
- planning value $p^*$
:::
:::::

------------------------------------------------------------------------

## 5.6.1 Sample Size for a Mean When $\sigma$ is Known

::: defbox
For a confidence interval on a population mean with known $\sigma$:

$$
E = z_{1-\alpha/2}\frac{\sigma}{\sqrt{n}}
$$

Solving for $n$ gives:

$$
n =
\left(
\frac{z_{1-\alpha/2}\sigma}{E}
\right)^2
$$
:::

::: {.warnbox .fragment}
Always round up to the next whole number. Rounding down would make the
margin of error too large.
:::

------------------------------------------------------------------------

## Example 1: Sample Size for a Mean, Known $\sigma$ {.scrollable}

:::::: two-col-wide
:::: col
::: taskbox
[Example]{.label} A lab measures chemical concentration. Repeated
measurements have known standard deviation:

$$
\sigma = 0.0068
$$

The lab wants a 95% confidence interval with margin of error at most:

$$
E = 0.005
$$

How many repeated measurements are needed?

**(a)** Identify $z^*$.

**(b)** Compute the required sample size.

**(c)** Explain why we round up.
:::
::::

::: col
::: {.examplebox}
Use:

$$
n =
\left(
\frac{z^*\sigma}{E}
\right)^2
$$

For 95% confidence:

$$
z^* \approx 1.96
$$
:::

::: {.thinkbox}
If the calculation gives 7.1, using 7 would not be enough. We must round
up to 8.
:::
:::
::::::

------------------------------------------------------------------------

## Sample Size Mean Example 1 — Check in R

```{webr}
#| autorun: false

sigma <- 0.0068
E <- 0.005
conf <- 0.95

alpha <- 1 - conf
z_star <- qnorm(1 - alpha / 2)

n_raw <- (z_star * sigma / E)^2
n_required <- ceiling(n_raw)

c(
  z_star = z_star,
  n_raw = n_raw,
  n_required = n_required
)
```

::: {.small-text}
The required sample size is $n=8$.
:::

------------------------------------------------------------------------

## Example 2: Sample Size for a Mean, Known $\sigma$ {.scrollable}

:::::: two-col-wide
:::: col
::: taskbox
[Example]{.label} A researcher wants to estimate average weekly spending.
A previous study suggests the population standard deviation is:

$$
\sigma = 22.5
$$

The researcher wants a 95% confidence interval with margin of error:

$$
E = 2
$$

What sample size is required?

**(a)** Write the sample size formula.

**(b)** Substitute the values.

**(c)** Round up and interpret.
:::
::::

::: col
::: {.examplebox}
The required sample size increases when:

- confidence level increases
- desired margin of error decreases
- population standard deviation increases
:::

::: {.thinkbox}
Smaller margin of error means we are demanding more precision, so the
sample size must increase.
:::
:::
::::::

------------------------------------------------------------------------

## Sample Size Mean Example 2 — Check in R

```{webr}
#| autorun: false

sigma <- 22.5
E <- 2
conf <- 0.95

alpha <- 1 - conf
z_star <- qnorm(1 - alpha / 2)

n_raw <- (z_star * sigma / E)^2
n_required <- ceiling(n_raw)

c(
  z_star = z_star,
  n_raw = n_raw,
  n_required = n_required
)
```

::: {.small-text}
The required sample size is $n=487$.
:::

------------------------------------------------------------------------

## Sample Size for a Mean When $\sigma$ is Unknown

::: defbox
When $\sigma$ is unknown, we use a preliminary estimate $s$ from a pilot
study or previous data.

A practical formula is:

$$
n =
\left(
\frac{t_{1-\alpha/2,\;n-1}s}{E}
\right)^2
$$

but the $t$ critical value depends on $n$, the quantity we are trying to
find.
:::

::: {.thinkbox .fragment}
This is why an iterative approach is used:

1. Start with the $z$ approximation.
2. Use that $n$ to get a $t$ critical value.
3. Recalculate $n$.
4. Repeat until $n$ stabilizes.
5. Round up.
:::

------------------------------------------------------------------------

## Example 1: Unknown $\sigma$, Iterative Sample Size {.scrollable}

:::::: two-col-wide
:::: col
::: taskbox
[Example]{.label} A pilot study estimates the standard deviation of
response time as:

$$
s = 12
$$

A researcher wants a 95% confidence interval with margin of error:

$$
E = 5
$$

Find the required sample size using the iterative $t$ approach.

**(a)** Start with the $z$ approximation.

**(b)** Update using the $t$ critical value.

**(c)** Stop when the sample size stabilizes.
:::
::::

::: col
::: {.examplebox}
The unknown-$\sigma$ case is more realistic, but it is also more awkward
because the critical value depends on the final sample size.
:::

::: {.thinkbox}
The answer should be rounded up because the actual sample size must be a
whole number and must meet the desired precision.
:::
:::
::::::

------------------------------------------------------------------------

## Unknown $\sigma$ Example 1 — Check in R

```{webr}
#| autorun: false

s <- 12
E <- 5
conf <- 0.95
alpha <- 1 - conf

# Start with z approximation
n_current <- ceiling((qnorm(1 - alpha / 2) * s / E)^2)

history <- data.frame(
  step = 0,
  critical_value = qnorm(1 - alpha / 2),
  n = n_current
)

for (i in 1:10) {
  t_star <- qt(1 - alpha / 2, df = n_current - 1)
  n_next <- ceiling((t_star * s / E)^2)

  history <- rbind(
    history,
    data.frame(
      step = i,
      critical_value = t_star,
      n = n_next
    )
  )

  if (n_next == n_current) {
    break
  }

  n_current <- n_next
}

history
```

::: {.small-text}
The sample size stabilizes at $n=25$.
:::

------------------------------------------------------------------------

## Example 2: Unknown $\sigma$, Iterative Sample Size {.scrollable}

:::::: two-col-wide
:::: col
::: taskbox
[Example]{.label} A pilot study estimates the standard deviation of
monthly textbook spending as:

$$
s = 18
$$

A 90% confidence interval is desired with margin of error:

$$
E = 4
$$

Find the required sample size using the iterative $t$ approach.

**(a)** Start with the $z$ approximation.

**(b)** Recalculate using the $t$ critical value.

**(c)** Report the required sample size.
:::
::::

::: col
::: {.examplebox}
For 90% confidence, the critical value is smaller than for 95% confidence,
but the desired margin of error is also quite small.
:::

::: {.thinkbox}
The final answer is the first sample size where the iteration stops
changing.
:::
:::
::::::

------------------------------------------------------------------------

## Unknown $\sigma$ Example 2 — Check in R

```{webr}
#| autorun: false

s <- 18
E <- 4
conf <- 0.90
alpha <- 1 - conf

n_current <- ceiling((qnorm(1 - alpha / 2) * s / E)^2)

history <- data.frame(
  step = 0,
  critical_value = qnorm(1 - alpha / 2),
  n = n_current
)

for (i in 1:10) {
  t_star <- qt(1 - alpha / 2, df = n_current - 1)
  n_next <- ceiling((t_star * s / E)^2)

  history <- rbind(
    history,
    data.frame(
      step = i,
      critical_value = t_star,
      n = n_next
    )
  )

  if (n_next == n_current) {
    break
  }

  n_current <- n_next
}

history
```

::: {.small-text}
The sample size stabilizes at $n=57$.
:::

------------------------------------------------------------------------

## Visual: Margin of Error Shrinks as $n$ Grows

```{r}
#| echo: false
#| warning: false
#| message: false
#| out-width: "100%"

sigma <- 22.5
z_star <- qnorm(0.975)

n_vals <- seq(25, 600, by = 5)
me_vals <- z_star * sigma / sqrt(n_vals)

df_me <- data.frame(
  n = n_vals,
  margin_error = me_vals
)

p <- plot_ly(
  data = df_me,
  x = ~n,
  y = ~margin_error,
  type = "scatter",
  mode = "lines",
  line = list(color = plot_cols$main, width = 3.6),
  text = ~paste0(
    "<b>Sample size:</b> ", n,
    "<br><b>Margin of error:</b> ", round(margin_error, 2)
  ),
  hovertemplate = "%{text}<extra></extra>",
  showlegend = FALSE
)

shapes <- list(
  list(
    type = "line",
    x0 = 25,
    x1 = 600,
    y0 = 2,
    y1 = 2,
    line = list(color = plot_cols$accent, width = 2.5, dash = "dash")
  )
)

plotly_slide_layout(
  p,
  title = "Margin of error decreases as sample size increases",
  x_title = "Sample size n",
  y_title = "Margin of error",
  height = 390,
  x_range = c(25, 600),
  y_range = c(0, 10),
  shapes = shapes
)
```

::: {.thinkbox}
The dashed line marks the target margin of error $E=2$. To cut the margin
of error in half, we need about four times as much data.
:::

------------------------------------------------------------------------

## 5.6.2 Sample Size for a Proportion

::: defbox
For a confidence interval on a population proportion, the margin of error
is approximately:

$$
E =
z_{1-\alpha/2}
\sqrt{
\frac{p^*(1-p^*)}{n}
}
$$

Solving for $n$ gives:

$$
n =
\left(
\frac{z_{1-\alpha/2}}{E}
\right)^2
p^*(1-p^*)
$$
:::

::: {.thinkbox .fragment}
The value $p^*$ is a planning value. If no prior information is
available, use $p^*=0.5$ because it gives the most conservative sample
size.
:::

------------------------------------------------------------------------

## Example 1: Sample Size for a Proportion {.scrollable}

:::::: two-col-wide
:::: col
::: taskbox
[Example]{.label} A city wants to estimate the proportion of voters who
support a candidate.

The city wants a 95% confidence interval with margin of error:

$$
E=0.03
$$

No prior estimate of the population proportion is available.

How large should the sample be?

**(a)** What planning value should be used?

**(b)** Calculate the required sample size.

**(c)** Explain why we round up.
:::
::::

::: col
::: {.examplebox}
With no prior information, use:

$$
p^*=0.5
$$

because $p^*(1-p^*)$ is largest when $p^*=0.5$.
:::

::: {.thinkbox}
This choice is conservative: it protects us from underestimating the
needed sample size.
:::
:::
::::::

------------------------------------------------------------------------

## Sample Size Proportion Example 1 — Check in R

```{webr}
#| autorun: false

p_star <- 0.5
E <- 0.03
conf <- 0.95

alpha <- 1 - conf
z_star <- qnorm(1 - alpha / 2)

n_raw <- (z_star / E)^2 * p_star * (1 - p_star)
n_required <- ceiling(n_raw)

c(
  z_star = z_star,
  p_star = p_star,
  n_raw = n_raw,
  n_required = n_required
)
```

::: {.small-text}
The required sample size is $n=1068$.
:::

------------------------------------------------------------------------

## Example 2: Sample Size for a Proportion {.scrollable}

:::::: two-col-wide
:::: col
::: taskbox
[Example]{.label} A consumer advocacy group wants to estimate the
proportion of consumers who bought a new phone and were happy with their
purchase.

They want a 90% confidence interval with margin of error:

$$
E=0.02
$$

No prior estimate is available.

How large should the sample be?

**(a)** Choose a planning value.

**(b)** Use $z^*=1.645$ for 90% confidence.

**(c)** Calculate and round up.
:::
::::

::: col
::: {.examplebox}
No prior estimate is given, so use:

$$
p^*=0.5
$$

For 90% confidence, a common rounded critical value is:

$$
z^*=1.645
$$
:::

::: {.thinkbox}
The margin of error is only 2%, so the required sample size will be
large.
:::
:::
::::::

------------------------------------------------------------------------

## Sample Size Proportion Example 2 — Check in R

```{webr}
#| autorun: false

p_star <- 0.5
E <- 0.02

# Using the common rounded 90% critical value
z_star <- 1.645

n_raw <- (z_star / E)^2 * p_star * (1 - p_star)
n_required <- ceiling(n_raw)

c(
  z_star = z_star,
  p_star = p_star,
  n_raw = n_raw,
  n_required = n_required
)
```

::: {.small-text}
Using $z^*=1.645$, the required sample size is $n=1692$.
:::

------------------------------------------------------------------------

##  {.quiz-question}

::: {.callout-note icon="false"}
## Question

When planning a confidence interval for a population proportion, no prior
estimate of $p$ is available. What value should we use for $p^*$?
:::

-   $p^*=0$
-   $p^*=0.25$
-   [$p^*=0.5$]{.correct
    data-explanation="When no prior estimate is available, p*=0.5 is used because p*(1-p*) is maximized at 0.5. This gives the most conservative required sample size."}
-   $p^*=1$